Structured Versus Decorated Cospans

### 2020-12-06 cospans.tex
began work on gray-boxing in 5.4

\begin{thm}
\label{thm:functoriality}
Given two categories $\A$ and $\A'$ with finite colimits, two symmetric lax monoidal pseudofunctors $F \maps \A \to \Cat$ and $F' \maps \A' \to \Cat$, a finite colimit preserving functor $H \maps \A \to \A'$, a symmetric lax monoidal pseudofunctor $E \maps \Cat \to \Cat$ and a 2-isomorphism $\theta \maps EF \Rightarrow F' H$ as in the following diagram,
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\A$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\mathsf{A'}$};
\node (C) at (0,-1) {$\A'$};
\node (D) at (1,-1) {$\Cat$};
\node (E) at (0.5,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$H$} (C)
(B) edge node[right]{$E$} (D)
(C) edge node[above]{$F'$} (D);
\end{tikzpicture}
\] 
the triple $(H,E,\theta)$ induces a symmetric monoidal double functor $\lH \maps F\lCsp \to F'\lCsp$ as defined above.
\end{thm}

@@

In chemistry, population biology, epidemiology and other fields, modelers often use `Petri nets with rates', where the transitions are labelled with positive real numbers called `rate constants' \cite{Haas,Koch,Wilkinson}.   From any Petri net with rates one can systematically construct a dynamical system.  Mathematical chemists have proved deep theorems relating the topology of Petri nets with rates to the qualitative behavior of their dynamical systems \cite{CTF}.

@@

Now let us turn to the details.  A \define{Petri net with rates} is a Petri net $s,t \maps T \to \N[S]$ together with a function $r \maps T \to (0,\infty)$ assigning to each transition $\tau \in T$ a positive real number called its \define{rate constant}.  There is a category $\Petri_r$ whose objects are Petri nets with rates, where a morphism from 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
 to 
 \[   \xymatrix{ (0,\infty) & T' \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S'] }\]
is a morphism of the underlying Petri nets such that the following diagram also commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,-0.5) {$(0,\infty)$};
\node (G) at (4,0) {$T$};
\node (G') at (4,-1) {$T'$};
\path[->,font=\scriptsize,>=angle 90]
(G) edge node[right]{$f$} (G')
(G) edge node[above]{$r$} (E)
(G') edge node[below]{$r'$} (E);
\end{tikzpicture}
\]
There is a functor $R \maps \Petri_r \to \Set$ sending any Petri net with rates to its set of places, and this has a left adjoint $L \maps \Set \to \Petri_r$ sending any set $S$ to the Petri net with $S$ as its set of species and no transitions \cite[Lemma 6.18]{BC}.   Since $\Petri_r$ has finite colimits \cite[Lemma 6.19]{BC}, it follows that there is a symmetric monoidal double category ${}_L \lCsp(\Petri_r)$ where:
\begin{itemize}
\item objects are finite sets,
\item vertical 1-morphisms are functions,
\item horizontal 1-cells are \define{open Petri nets with rates}:
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
\item 2-morphisms are diagrams in $\Petri_r$ of the form:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,0) {$L(X_1)$};
\node (F) at (5,0) {$L(Y_1)$};
\node (G) at (4,0) {$P_1$};
\node (E') at (3,-1) {$L(X_2)$};
\node (F') at (5,-1) {$L(Y_2)$.};
\node (G') at (4,-1) {$P_2$};
\path[->,font=\scriptsize,>=angle 90]
(F) edge node[above]{$O_1$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$I_1$} (G)
(E') edge node[above]{$I_2$} (G')
(F') edge node[above]{$O_2$} (G');
\end{tikzpicture}
\]
\end{itemize}

@@

We now show that while open dynamical systems are morphisms in a decorated cospan category, this does not arise from a structured cospan category via \cref{thm:equiv}.  To do this, recall that in that theorem we start with the data required to build a decorated cospan category, namely a symmetric lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat,\times)$, and show that under certain conditions the resulting opfibration $U \maps \inta F \to \A$ has a left adjoint $L \maps \A \to \X$.   We then obtain an equivalence between decorated and structured cospan double categories, $F \lCsp \simeq {}_L \Csp(\X)$.   We now show that in the case at hand, where $F = D$, $U$ does \emph{not} have a left adjoint.  Thus, the conditions of \cref{thm:equiv} do not hold in this case.

@@

Even though for both \cref{thm:equiv} and the above described setting, the conditions required are only sufficient, they provide certain indication that under a suitable ``condition intersection'' in a sense, we could establish a functorial picture
\begin{displaymath}
\begin{tikzpicture}
\node [align=center,draw] (A) { Lax monoidal pseudofunctors \\ $(\A,+)\to(\Cat,\times)$};
\node[below right=1.5 and 1.5 of A,align=center,draw] (C)  { Symmetric monoidal \\double categories };
\node[below left=1 and .5  of A,draw] (D)  {Special opfibrations};
\node[below left=2 and 1 of A,draw] (E)  {Special ralis};
\node[below=3 of A,align=center,draw] (B)  {Finite colimit preserving \\ functors $A\to\X$};
\node[below=.01 of D] {\rotatebox{90}{$\simeq$}};
\path[->,font=\scriptsize]
(A) edge node[above,sloped]{$F\mapsto F\lCsp$}  (C)
(D) edge (A)
(E) edge (B)
(B) edge node[below,sloped]{$L\mapsto {}_{L}\lCsp$} (C);
\end{tikzpicture}
\end{displaymath}
with an actual natural isomorphism in the middle. This would imply that starting from an appropriate middle ground, employing the structured or the decorated cospan formalism is essentially the same. We leave such considerations for future work.

>>>>

\begin{thm}
\label{thm:functoriality}
Given two categories $\A$ and $\A'$ with finite colimits, two symmetric lax monoidal pseudofunctors $F \maps \A \to \Cat$ and $F' \maps \A' \to \Cat$, a finite colimit preserving functor $H \maps \A \to \A'$, a symmetric lax monoidal pseudofunctor $E \maps \Cat \to \Cat$ and a 2-isomorphism $\theta \maps EF \Rightarrow F' H$ as in the following diagram,
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\A$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\A'$};
\node (D) at (1,-1) {$\Cat$};
\node (E) at (0.5,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$H$} (C)
(B) edge node[right]{$E$} (D)
(C) edge node[above]{$F'$} (D);
\end{tikzpicture}
\] 
the triple $(H,E,\theta)$ induces a symmetric monoidal double functor $\lH \maps F\lCsp \to F'\lCsp$ as defined above.
\end{thm}

@@

In chemistry, population biology, epidemiology and other fields, modelers use `Petri nets with rates', where the transitions are labelled with positive real numbers called `rate constants' \cite{Haas,Koch,Wilkinson}.   From any Petri net with rates one can systematically construct a dynamical system.  Mathematical chemists have proved deep theorems relating the topology of Petri nets with rates to the qualitative behavior of their dynamical systems \cite{CTF}.

@@

Now we turn to the details.  A \define{Petri net with rates} is a Petri net $s,t \maps T \to \N[S]$ together with a function $r \maps T \to (0,\infty)$ assigning to each transition $\tau \in T$ a positive real number called its \define{rate constant}.  There is a category $\Petri_r$ whose objects are Petri nets with rates, where a morphism from 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
 to 
 \[   \xymatrix{ (0,\infty) & T' \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S'] }\]
is a morphism of the underlying Petri nets such that the following diagram also commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,-0.5) {$(0,\infty)$};
\node (G) at (4,0) {$T$};
\node (G') at (4,-1) {$T'$};
\path[->,font=\scriptsize,>=angle 90]
(G) edge node[right]{$f$} (G')
(G) edge node[above]{$r$} (E)
(G') edge node[below]{$r'$} (E);
\end{tikzpicture}
\]
There is a functor $R \maps \Petri_r \to \Set$ sending any Petri net with rates to its set of places, and this has a left adjoint $L \maps \Set \to \Petri_r$ sending any set $S$ to the Petri net with $S$ as its set of species and no transitions \cite[Lemma 6.18]{BC}.   Since $\Petri_r$ has finite colimits \cite[Lemma 6.19]{BC}, it follows that there is a symmetric monoidal double category ${}_L \lCsp(\Petri_r)$ where:
\begin{itemize}
\item objects are finite sets,
\item vertical 1-morphisms are functions,
\item horizontal 1-cells are \define{open Petri nets with rates}:
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
\item 2-morphisms are diagrams in $\Petri_r$ of the form:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,0) {$L(X_1)$};
\node (F) at (5,0) {$L(Y_1)$};
\node (G) at (4,0) {$P_1$};
\node (E') at (3,-1) {$L(X_2)$};
\node (F') at (5,-1) {$L(Y_2)$.};
\node (G') at (4,-1) {$P_2$};
\path[->,font=\scriptsize,>=angle 90]
(F) edge node[above]{$O_1$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$I_1$} (G)
(E') edge node[above]{$I_2$} (G')
(F') edge node[above]{$O_2$} (G');
\end{tikzpicture}
\]
\end{itemize}

@@

Next, we can define a symmetric monoidal double functor
\[      \graysquare \maps F \lCsp \to D \lCsp \]
sending any open Petri net with rates to its corresponding open dynamical system.  This was already defined as a functor between categories by Pollard and the first author \cite[Section 7]{BP}, who called it `gray-boxing'.   To boost this result to the double category level we use \cref{thm:functoriality}, taking the square in that theorem as follows:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\Fin\Set$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\Fin\Set$};
\node (D) at (1,-1) {$\Cat$};
\node (E) at (0.5,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$1$} (C)
(B) edge node[right]{$E$} (D)
(C) edge node[above]{$D$} (D);
\end{tikzpicture}
\] 
%%% continue this....

We now show that decorated cospan double category $D \lCsp$ of open dynamical systems is not equivalent to a structured cospan double category via \cref{thm:equiv}.  Recall that in that theorem we start with the data required to build a decorated cospan category, namely a symmetric lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat,\times)$, and show that under certain conditions the resulting opfibration $U \maps \X = \inta F \to \A$ has a left adjoint $L \maps \A \to \X$.   We then obtain an equivalence between decorated and structured cospan double categories, $F \lCsp \simeq {}_L \Csp(\X)$.   We now show that in the case at hand, where $F = D$,  the opfibration $U$ does \emph{not} have a left adjoint.  Thus, the conditions of \cref{thm:equiv} do not hold in this case.

@@

Even though for both \cref{thm:equiv} and the above described setting, the conditions required are only sufficient, they provide certain indication that under a suitable ``condition intersection'' in a sense, we could establish a functorial picture
\begin{displaymath}
\begin{tikzpicture}
\node [align=center,draw] (A) { Lax monoidal pseudofunctors \\ $(\A,+)\to(\Cat,\times)$};
\node[below right=1 and 1.5 of A,align=center,draw] (C)  { Symmetric monoidal \\double categories };
\node[below left=1.05 and .5  of A,draw] (D)  {Special opfibrations};
\node[below left=2 and 1 of A,draw] (E)  {Special ralis};
\node[below=3 of A,align=center,draw] (B)  {Finite colimit preserving \\ functors $A\to\X$};
\node[below=0 of D] {\rotatebox{90}{$\simeq$}};
\path[->,font=\scriptsize]
(A) edge node[above,sloped]{$F\mapsto F\lCsp$}  (C)
(D) edge (A)
(E) edge (B)
(B) edge node[below,sloped]{$L\mapsto {}_{L}\lCsp(\X)$} (C);
\end{tikzpicture}
\end{displaymath}
with an actual natural isomorphism in the middle. This would imply that starting from an appropriate middle ground, employing the structured or the decorated cospan formalism is essentially the same. We leave such considerations for future work.
